Power simulation section

### 2023-09-07 docs/manuscript.Rmd
improve the manuscript and add some comments

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, echo = TRUE, eval = TRUE)
```

@@

<!-- General short intro paragraph, need for well-powered and analyzed experimental research -->
When planning experimental research it is essential to determine an appropriate sample size [@lakensSampleSizeJustification2022] to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data. However, the recent replication crisis in experimental research, such as in Human-Computer Interaction (HCI), has illustrated many challenges surrounding the reproducibility and reliability of findings. As a results, there is a growing need for rigorous statistical methodology and the adoption of well-powered experimental designs. While accessible software solutions exists for simple statistical models and experimental designs, many experimental researcher lack the skills and tools to conduct power analyses for more complex research designs using powerful generalized linear mixed models (GLMM). In the present case study, we provide a tutorial comprised of a concrete example for custom a priori power analyses using data simulations based on GLMMs.

@@

<!-- Some basics on statistical power -->
In empirical research relying on hypothesis testing, the most common strategy for determining an adequate sample size is based on statistical power. Statistical power is defined as the probability that a hypothesis test turns out significant when analyzing repeated samples from a population with a true effect of some pre-specified size. Less formally, power is also described as the probability that a hypothesis test accepts the alternative hypothesis $H_1$ if $H_1$ is indeed true. If the sample size (i.e., the number of participants or observations) used for data collection is insufficient to reliably detect the effects or relationships being investigated., this particular study would be considered "under-powered".

<!-- what are the resulting issues with under-powered studies? -->
Conducting under-powered research has many negative consequences. First, relying on under-powered experiments may yield inconclusive or misleading results hindering the accumulation of knowledge. Second, under-powered studies waste resources by consuming time, effort, and funding without delivering meaningful results, thereby missing valuable discoveries and opportunities for scientific advancement.

@@

<!-- define power analysis -->
A power analysis represents the act of calculating the statistical power for a given true effect and sample size. 
By running the calculation for a range of sample sizes "a priori" (i.e., beforehand), the required sample size can be determined so that researchers find an assumed true effect with the desired statistical power. 
Thereby, power analysis offers a valuable solution by allowing researchers to estimate the appropriate sample sizes required to achieve sufficient statistical power for results with high evidential value.

<!-- advantages of an a priori power analysis -->
Conducting a careful a priori power analysis helps researchers decide which experimental design and statistical models are both feasible and appropriate for answering their research questions and analyzing the data.
Also, while conducting a proper power analysis, one has to consider every aspect of the experimental design. This makes one think through everything carefully and might highlight statistical or design challenges before the start of an experiment. 
This gives researchers a great opportunity to improve the design. Especially in fast-moving research fields, where speed is often regarded as more important than rigor, adding a solid sample size calculation into the process can act as a safeguard for ensuring high-quality research.

<!-- external incentives to do a power analysis -->
Moreover, many journals and funding agencies now mandate the inclusion of power analysis as part of study protocols and grant proposals, recognizing their significance in ensuring robust and meaningful findings. By incorporating power analysis into research planning, researchers can enhance the replicability and credibility of their work, contributing to the advancement of the respective research field through the gained statistically-sound insights.

<!-- standard software is available for "simple" models and study designs -->
For simple statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups) user-friendly power analysis software is readily available [@faulStatisticalPowerAnalyses2009].

@@

<!-- Advantages of GLMMs over "simpler" models -->
As study designs become more complex, researchers require more sophisticated statistical models to capture the nuanced relationships and hierarchical structures introduced by their study designs [@yarkoniGeneralizabilityCrisis2022]. As a result, GLMMs (generalized) linear mixed models (GLMMs; also called multilevel models) are gaining increasing popularity in analyzing data from experimental research, such as in HCI, because they offer a flexible framework for analyzing data with non-normal and categorical outcomes while accounting for both fixed and random effects [@kapteinUsingGeneralizedLinear2016; @fahrmeirRegressionModelsMethods2021].

@@

<!-- what are the existing option for a priori power analysis for GLMMs / where are existing tutorials limited -->
Currently available software packages for power simulations do not include GLMMs, making it necessary to build data simulations tailored specifically to the study design [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022]. 
A number of tutorials have been published describing how to perform such power simulation for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. 
However, most of these tutorials only cover linear mixed models (LMMs) and focus on the most common designs (e.g., INSERT EXAMPLE). This narrow focus provides little guidance for researchers faced with more complex study designs (e.g., INSERT EXAMPLE), which are very common [@brownIntroductionLinearMixedEffects2021]. 
The necessary assumptions for power simulations in GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the error structure. The distributional assumption specifies the family of distributions for the outcome variable, such as Gaussian, Poisson, or binomial. Assumptions about the random effects include the assumption of normality and the covariance structure among the random effects. Additionally, assumptions about the error structure, such as independence or correlation, must be specified. Interpreting these assumptions entails understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Here, existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, rendering estimation a practical challenge [@maxwellSampleSizePlanning2008]. 
<!-- Summary / Wrap-Up, pick up main arguments and what the paper provides, keep this short! -->
Based on need for well-powered experimental research using GLMMs and the lack of tools to conduct corresponding power analyses, in this tutorial paper, we present a case study that serves as a practical demonstration of how to perform a data simulation for a priori power simulations of such study designs. Thereby, we aim to equip researchers with the tools needed to simulate their own data and determine appropriate sample sizes with GLMMs.

@@

In this section, we outline the steps for performing data simulation and a priori power analysis for GLMMs, using a case study based on a specific experimental study design from the area of human-AI interaction research.

@@

<!-- describe the study idea and design -->
In the present case study, we simulate the data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system is to be evaluated. 
The goal is to understand how AI advice influences medical decision-making. Participants, radiologists (task experts) and students/interns (non-task experts), review head CT (computer tomography) scans to assess the presence of bleeding, more specifically, intracranial haemorrhage (ICH). 
To support their decision-making, in two experimental conditions, an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either either correct (80% of cases) or incorrect (20%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate. Each participant undergoes all experimental conditions.
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans, (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice, and (c) different presentations of AI advice influence performance.
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible. 
The goal of the power simulation is to determine how many task experts and non-experts to recruit to achieve sufficient statistical power in our experiment.

@@

<!-- describe the lme4 package and how it works-->
In our case study, we use the lme4 R package [@batesFittingLinearMixedEffects2015]. The lme4 package is a state-of-the-art tool for fitting frequentist GLMMs.[^1]

@@

However, the model parameterization used by the lme4 package is quite technical, making it difficult for applied researchers to determine whether their specified population model (i.e., the theoretical model that describes the underlying data generation process for a specific population of interest) implies plausible associations in their simulated data. Therefore in this tutorial, we simulate data for GLMMs from first principles (i.e., starting from the most basic and fundamental concepts and building up from there to create synthetic data) to assist applied researchers in better understanding all model assumptions and then use lme4 to analyze the simulated data sets.[^2]

@@

<!-- deep dive into our specific example -->
In a GLMM, the expected value of the dependent variable Y conditioned on the vector of predictor variables $\mathbf{X}$ and random effects $\mathbf{U}$, transformed by a link function $g()$ is modelled as a linear combination $\eta$ of the predictor variables $\mathbf{X}$, the random effects $\mathbf{U}$ and the model parameters $\mathbf{\beta}$ [@fahrmeirRegressionModelsMethods2021]. $$
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
$$ Equivalently, the conditional expected value is modelled as the linear combination $\eta$, transformed by the inverse link function $g^{-1}()$. $$
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = g^{-1}(\eta)
$$ If the dependent variable (i.e., performance) $Y$ is a binary variable with values $0$ or $1$, the conditional expected value is equivalent to the probability: $$
P_{si} = P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
$$ In our case study, $P_{si}$ is the conditional probability that subject (i.e., participant) $s$ gives the correct response to item (i.e., CT scan) $i$.

In such a setting, we model the probability as $$
P_{si} = inverse\_logit(\eta_{si})
$$ with the inverse-logit link $g^{-1}(\eta_{si}) = inverse\_logit(\eta_{si}) = \frac{exp(\eta_{si})}{1 + exp(\eta_{si})}$ or equivalently $$
logit(P_{si}) = \eta_{si}
$$ with the logit link $g(P_{si}) = logit(P_{si}) = ln (\frac{P_{si}}{1 - P_{si}})$.

@@

In our case study, each subject (`n_subjects` in total) is assumed to respond to each item (i.e., CT scan) (`n_items` in total). Thus the `add_random` command creates a fully-crossed `data.frame` with `n_subjects` $\times$ `n_items` rows. We add a between-subject effect with the `add_between` command, simulating that about $25\%$ of subjects are experts. The next two lines simulate that in about $\frac{2}{3}$ of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about $80\%$ of cases (the variable `advice_correct` is always 0 when no advice is presented). Next, we simulate one random effect for each subject (`u0s`) and for each item (`u0i`). As assumed by standard GLMMs, the `add_ranef` function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user. With all design variables done, we are ready to simulate our model equation as outlined in equation X. The linear predictor variable `linpred` ($\eta$ in the GLMM model equations) combines the predictor variables, random effects and model parameters as assumed by our model. We then transform the linear predictor with the inverse-link function to compute `y_prob`, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with `plogis` and the logit link with `qlogis`). In the final step, we simulate the binary dependent variable `y_bin` by -- for each trial -- drawing from a Bernoulli distribution with success probability `y_prob`.

@@

The lme4 package uses a special syntax for model specification. Our assumed GLMM is represented by the formula:

@@

In lme4, a GLMM is fitted with the `glmer` function. By setting `family =  "binomial"`, we request a binomial GLMM appropriate for our binary dependent variable `y_bin` (the binomial GLMM uses the canonical logit link by default), which is defined as an accurate vs. inaccurate diagnosis.

@@

We can inspect the estimated model parameters with the `summary` command:

@@

The output shows the estimates for all model parameters: the `Estimate` column in the `Fixed effects` table contains the estimates for the $\beta$ parameters, while the `Std.Dev.` column in the `Random effects` table contains the estimates for $\sigma_S$ and $\sigma_I$.

Unfortunately, the model parameters in a binomial GLMM are hard to interpret because 1) the $\beta$ parameters are connected to the modelled probability via the non-linear inverse-logit link, and 2) we also have to consider the random effects. The most simple interpretation works by imagining a subject with average ability ($u_{0s} = 0$) responding to an item (i.e., CT scan) with average difficulty ($u_{0i} = 0$). Then the model implied probability that such a person solves such an item accurately is given by: $$
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= inverse\_logit[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$

@@

Due to this complicated relationship, we argue not to focus too much on interpreting single model parameters when working with GLMMs. Instead, it can be more intuitive to consider the model-implied distribution of the dependent variable for each experimental condition across all subjects and items.

@@

The `Fixed effects` table in Figure X also includes p-values for hypothesis tests with null hypotheses of the style $H_0: \beta = 0$. However, for many research questions of interest, we are not interested in these two-sided tests referring to only one parameter.

@@

Although the derivations above are straightforward, it is important not to misinterpret their implications: In binomial GLMMs, the average probability to solve an item (averaged across persons of varying ability and items of varying difficulty) is **not** equal to the probability that a person with average ability solves an item with average difficulty. For example, we determined the $\beta$ parameters in a way that corresponds to a desired probability of $0.95$, that an expert with average ability solves an item with average difficulty when presented with correct advice. However, even if the model were true, we would not observe this probability if we estimated the probability in a group of experts responding to items presented with correct advice from a big sample of subjects drawn from their natural distribution of ability and items drawn from their natural distribution of difficulty. This implies that one must be careful when specifying parameter values based on previous studies or pilot data. <!-- SG: What does that mean for someone who wants to do this then?-->

The inequality of conditional and marginal effects in GLMMs (i.e., how predictor variables affect the response within groups versus their overall impact across all groups, which can differ due to random effects) [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult. However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge. Then we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

@@

The final plot demonstrates that these plots are also useful for spotting standard deviations that are specified too high. For this example, we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are usually close to either 0 or 1, which is not a plausible assumption. <!-- SG: Maybe explain why this is not plausible.-->

@@

With all these considerations addressed, we are finally ready to perform a power analysis. Wrapping the `simulate` function already introduced earlier, the helper function `sim_and_analyse` performs all previous steps (simulate a dataset, fit a GLMM, compute p-values) in a single command.

@@

```{r, message=FALSE, eval=FALSE}
library(future)
plan("multisession", workers = 4)
set.seed(2)
```

The next code chunk specifies a simulation grid with different settings for both the number of subjects (`n_subjects`) and the number of items (`n_items`), each combination being repeated `rep` times. We chose 300 repetitions for the data simulation at hand as it strikes a balance between achieving a robust statistical estimate and remaining computationally feasible.

@@

The result of the computation is a data.frame that contains the p-values of all tested hypotheses for each simulated dataset.

@@

[^3]: This code is heavily inspired by the "Mixed Design Simulation" vignette of the faux package at <https://debruine.github.io/faux/articles/sim_mixed.html>.

```{r, eval=FALSE}
library(binom)
alpha <- 0.05
power <- sim_design %>%
  group_by(n_subjects, n_items) %>% 
  summarise(power = mean(p_H01 < alpha & p_H02 < alpha & 
                           p_H03 < alpha & p_H04 < alpha), 
    n_sig = sum(p_H01 < alpha & p_H02 < alpha & 
                  p_H03 < alpha & p_H04 < alpha),
    n = n(),
    ci.lwr = binom.confint(n_sig, n, method = "wilson")$lower,
    ci.upr = binom.confint(n_sig, n, method = "wilson")$upper, 
    .groups = "drop")
power %>%
  mutate(across(c(n_subjects, n_items), factor)) %>%
  ggplot(aes(n_subjects, n_items, fill = power)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                power, ci.lwr, ci.upr)), 
    color = "white", size = 6) +
  scale_fill_viridis_c(limits = c(0, 1))
```

>>>>

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, echo = TRUE, eval = FALSE)
```

@@

<!-- General short intro paragraph, need for well-powered and analyzed experimental research -->
When planning experimental research it is essential to determine an appropriate sample size [@lakensSampleSizeJustification2022] to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data. However, the recent replication crisis in experimental research, such as in Human-Computer Interaction (HCI), has illustrated many challenges surrounding the reproducibility and reliability of findings. <!-- FP: do we need a reference here? --> As a results, there is a growing need for rigorous statistical methodology and the adoption of well-powered experimental designs. While accessible software solutions exists for simple statistical models and experimental designs, many experimental researchers lack the skills and tools to conduct power analyses for more complex research designs using powerful <!-- FP: that can be confused with statistical power. Alternative: ... using the flexible GLMM framework --> generalized linear mixed models (GLMM). In the present case study, we provide a tutorial comprised of a concrete example for custom a priori power analyses using data simulations based on GLMMs.

@@

<!-- Some basics on statistical power -->
In empirical research relying on hypothesis testing, the most common strategy for determining an adequate sample size is based on statistical power <!-- FP:  cite lakens again here? -->. Statistical power is defined as the probability that a hypothesis test turns out significant <!-- FP: ... has a significant p-value? to make sure that we mean the statistical/technical meaning of significant  --> when analyzing repeated samples from a population with a true effect of some pre-specified size. Less formally, power is also described as the probability that a hypothesis test accepts the alternative hypothesis $H_1$ if $H_1$ is indeed true. If the sample size (i.e., the number of participants and/or stimuli) used for data collection is insufficient to reliably detect <!-- FP: alternative to reliably detect: detect the effects with high probability --> the effects or relationships being investigated, this particular study would be considered "under-powered".

<!-- what are the resulting issues with under-powered studies? -->
Conducting under-powered research has many negative consequences. First, relying on under-powered experiments may yield inconclusive (if low power is discussed by the researchers) or misleading (if low power is ignored) results, hindering the accumulation of knowledge. Second, under-powered studies waste resources by consuming time, effort, and funding without delivering meaningful results, thereby missing valuable discoveries and opportunities for scientific advancement <!-- FP: I would cut everything after funding  -->.

@@

<!-- define power analysis -->
A power analysis represents the act of calculating the statistical power for a given true effect and sample size. 
By running the calculation for a range of sample sizes "a priori" (i.e., before data collection), the required sample size can be determined so that researchers find an assumed true effect with the desired statistical power. 
Thereby, power analysis offers a valuable contribution to the research process by allowing researchers to estimate the appropriate sample sizes required to achieve sufficient statistical power for results with high evidential value.

<!-- advantages of an a priori power analysis -->
Conducting a careful a priori power analysis helps researchers decide which experimental design and statistical models are both feasible and appropriate for analyzing the data and answering their research questions.
Also, while conducting a proper power analysis, one has to consider every aspect of the experimental design and will possibly notice statistical or design challenges before the start of an experiment. 
This gives researchers a great opportunity to improve the design. Especially in fast-moving research fields, adding a solid sample size calculation into the process can act as a safeguard for ensuring high-quality research.

<!-- external incentives to do a power analysis -->
Moreover, many journals and funding agencies now mandate the inclusion of power analysis as part of study protocols and grant proposals, recognizing its significance in ensuring robust and meaningful findings. By incorporating power analysis into research planning, researchers can enhance the replicability and credibility of their work, contributing to the advancement of the respective research field through more robust insights. <!-- FP: this last sentence seems like a repetition  -->

<!-- standard software is available for "simple" models and study designs -->
For simple statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups) user-friendly software for power analysis is readily available [@faulStatisticalPowerAnalyses2009].

@@

<!-- Advantages of GLMMs over "simpler" models -->
As study designs become more complex, researchers require more sophisticated statistical models to capture the nuanced relationships and hierarchical structures introduced by their study designs [@yarkoniGeneralizabilityCrisis2022]. GLMMs (also called multilevel models) are gaining increasing popularity in analyzing data from experimental research, such as in HCI, because they offer a flexible framework for analyzing data with non-normal and categorical outcomes while accounting for both fixed and random effects [@kapteinUsingGeneralizedLinear2016; @fahrmeirRegressionModelsMethods2021].

@@

<!-- what are the existing option for a priori power analysis for GLMMs / where are existing tutorials limited -->
Currently available software packages for power simulations <!-- FP: because you removed the sentence on the difference between analytic and simulation based power analysis, it is now not clear why from now on we always talk about power simulation  --> do not include GLMMs, making it necessary to build data simulations tailored specifically to the study design [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022]. 
A number of tutorials have been published describing how to perform such power simulation for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. 
However, most of these tutorials only cover linear mixed models (LMMs) and focus on the most common designs (e.g., INSERT EXAMPLE). This narrow focus provides little guidance for researchers faced with more complex study designs (e.g., INSERT EXAMPLE), which are very common [@brownIntroductionLinearMixedEffects2021]. 
The necessary assumptions for power simulations in GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the error structure. The distributional assumption specifies the family of distributions for the outcome variable, such as Gaussian, Poisson, or binomial. Assumptions about the random effects include the assumption of normality and the covariance structure among the random effects. Additionally, assumptions about the error structure, such as independence or correlation, must be specified. Interpreting these assumptions entails understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Here, existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, rendering estimation <!-- FP: why estimation? more like specifying assumptions? --> a practical challenge [@maxwellSampleSizePlanning2008].
<!-- Summary / Wrap-Up, pick up main arguments and what the paper provides, keep this short! -->
Based on need for well-powered experimental research using GLMMs and the lack of tools to conduct corresponding power analyses, in this tutorial paper, we present a case study that serves as a practical demonstration of how to perform a data simulation and an a priori power simulation for a more complex study design. Thereby, we aim to equip researchers with the tools needed to simulate their own data and determine appropriate sample sizes for their own research.

@@

In this section, we outline the steps for performing data simulation and a priori power analysis for GLMMs, using a case study based on a specific experimental study design from the area of human-AI <!-- FP: this is the first time AI appears; probably have to define the AI abbreviation --> interaction research.

@@

<!-- describe the study idea and design -->
In the present case study, we simulate the data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system is to be evaluated. 
The goal is to understand how AI advice influences medical decision-making. Participants, radiologists (task experts) and students/interns (non-task experts) <!-- FP: during our work on the paper, I have heard two different stories of who the non experts are; either students/interns or full doctors that where not radiologists. Although this is only a data simulation example, it is nonetheless important here because the assumed effect sizes will probably differ! So who did you imagine when choosing the effect sizes?  -->, review head computer tomography (CT) scans to assess the presence of bleeding, more specifically, intracranial haemorrhage (ICH). 
To support their decision-making, in two experimental conditions, an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either correct (80% of cases) or incorrect (20%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate. Each participant undergoes all experimental conditions.
<!-- FP: I would remove the phrase experimental condition from this section. I think this was newly introduced here and does not match well with the experiment and the model as described later  -->
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans, (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice, and (c) different presentations of AI advice influence performance.
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible. 
The goal of the power simulation is to determine how many task experts and non-experts must be recruited to achieve sufficient statistical power in the planned experiment.

@@

<!-- describe the lme4 package and how it works-->
In our case study, we use the lme4 R package [@batesFittingLinearMixedEffects2015] which is a state-of-the-art tool for fitting frequentist GLMMs.[^1]

@@

However, the model parameterization used by the lme4 package is quite technical, making it difficult for applied researchers to determine whether their specified population model (i.e., the theoretical model that describes the underlying data generation process for a specific population of interest) implies plausible associations in their simulated data. Therefore in this tutorial, we simulate data for GLMMs from first principles (i.e., starting from the most basic and fundamental concepts and building up from there to create synthetic data <!-- FP: ALTERNATIVE: creating synthetic data step by step instead of using black box functions -->) to assist applied researchers in better understanding all model assumptions and then use lme4 to analyze the simulated data sets.[^2]

@@

<!-- deep dive into our specific example -->
In a GLMM, the expected value of the dependent variable Y conditioned on the vector of predictor variables $\mathbf{X}$ and random effects $\mathbf{U}$, transformed by a link function $g()$ is modeled as a linear combination $\eta$ of the predictor variables $\mathbf{X}$, the random effects $\mathbf{U}$ and the model parameters $\mathbf{\beta}$ [@fahrmeirRegressionModelsMethods2021]. $$
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
$$ Equivalently, the conditional expected value is modeled as the linear combination $\eta$, transformed by the inverse link function $g^{-1}()$. $$
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = g^{-1}(\eta)
$$ If the dependent variable (i.e., performance <!-- FP: is performance really the best option to describe our DV?  -->) $Y$ is a binary variable with values $0$ or $1$, the conditional expected value is equivalent to the probability: $$
P_{si} := P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
$$ In our case study, $P_{si}$ is the conditional probability that subject (i.e., participant <!-- FP: can't we just use subject instead of participant everywhere in the manuscript, i.e. also in the intro and when describing the experiment? I thought we had already agreed on this when deciding on how to write the equations and I do not want to change all the s subscripts now  -->) $s$ gives the correct response to item (i.e., CT scan) $i$.

In such a setting, we model this probability as $$
P_{si} = inverse\_logit(\eta_{si})
$$ with the inverse-logit link $g^{-1}(\eta_{si}) = inverse\_logit(\eta_{si}) = \frac{exp(\eta_{si})}{1 + exp(\eta_{si})}$ or equivalently $$
logit(P_{si}) = \eta_{si}
$$ with the logit link $g(P_{si}) = logit(P_{si}) = ln (\frac{P_{si}}{1 - P_{si}})$.

@@

In the first six lines of the function definition, we set some default parameter values (which we will explain in a later section) and load the packages we use to manipulate and simulate data. 
In our case study, each subject (`n_subjects` in total) is assumed to respond to each item (i.e., CT scan; `n_items` in total). Thus the `add_random` command creates a fully-crossed `data.frame` with `n_subjects` $\times$ `n_items` rows. We add a between-subject effect with the `add_between` command, simulating that about $25\%$ of subjects are experts. The next two lines simulate that in about $\frac{2}{3}$ of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about $80\%$ of cases (the variable `advice_correct` is always 0 when no advice is presented). Next, we simulate one random effect for each subject (`u0s`) and for each item (`u0i`). As assumed by standard GLMMs, the `add_ranef` function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user. With all design variables done, we are ready to simulate our model equation as outlined in equation X. The linear predictor variable `linpred` ($\eta$ in the GLMM model equations) combines the predictor variables, random effects and model parameters as assumed by our model. We then transform the linear predictor with the inverse-link function to compute `y_prob`, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with `plogis` and the logit link with `qlogis`). In the final step, we simulate the binary dependent variable `y_bin` (i.e., whether the subject chooses the correct diagnosis for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability `y_prob`.

@@

The lme4 package uses a special syntax for model specification. Our specific GLMM is represented by the formula:

@@

In lme4, a GLMM is fitted with the `glmer` function. By setting `family =  "binomial"`, we request a binomial GLMM appropriate for our binary dependent variable `y_bin` (the binomial GLMM uses the canonical logit link by default), which is defined as an accurate (1) vs. inaccurate (0) diagnosis.

@@

We can inspect the estimates for all model parameters with the `summary` command:

@@

In the output, the `Estimate` column in the `Fixed effects` table contains the estimates for the $\beta$ parameters, while the `Std.Dev.` column in the `Random effects` table contains the estimates for $\sigma_S$ and $\sigma_I$.

Unfortunately, the model parameters in a binomial GLMM are hard to interpret because 1) the $\beta$ parameters are connected to the modeled probability via the non-linear inverse-logit link, and 2) we also have to consider the random effects. The most simple interpretation works by imagining a subject with average ability ($u_{0s} = 0$) responding to an item (i.e., CT scan) with average difficulty ($u_{0i} = 0$). Then the model implied probability that such a person solves such an item accurately is given by: $$
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= inverse\_logit[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$

@@

Due to this complicated relationship, we argue not to focus too much on interpreting single model parameters when working with GLMMs. Instead, it can be more intuitive to consider model predictions and the model-implied distribution of the dependent variable for each experimental condition across all subjects and items.

@@

The `Fixed effects` table in Figure X also includes p-values for hypothesis tests with null hypotheses of the style $H_0: \beta = 0$. However, for many research questions of interest, we are not interested in these two-sided tests that refer to only a single parameter.

@@

Although the derivations above are straightforward, it is important not to misinterpret their implications: In binomial GLMMs, the average probability to solve an item (averaged across persons of varying ability and items of varying difficulty) is **not** equal to the probability that a person with average ability solves an item with average difficulty. The first perspective implies a so-called marginal interpretation while the second one implies a conditional interpretation. For example, we determined the $\beta$ parameters in a way that corresponds to a desired conditional probability of $0.95$, that an expert with average ability solves an item with average difficulty when presented with correct advice. However, even if the model were true, we would not observe this probability value if we estimated the marginal probability in a group of experts responding to items presented with correct advice from a big sample of subjects drawn from their natural distribution of ability and items drawn from their natural distribution of difficulty.

The inequality of conditional and marginal effects in GLMMs (i.e., how predictor variables affect the response within groups versus their overall impact across all groups, which can differ due to random effects) <!-- FP: I have tried to introduce contidional and marginal in the above section, so I would suggest to remove the bracket here --> [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult. One must be careful when specifying parameter values based on previous studies or pilot data that use the marginal interpretation (e.g., a pilot study providing an estimate of how often neurologists make an accurate diagnosis based on brain scans). <!-- SG: What does that mean for someone who wants to do this then? FP: is it more clear now? --> However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge. Then we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

@@

The final plot demonstrates that these plots are also useful for spotting standard deviations that are specified too high. For this example, we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are usually close to either 0 or 1, which is not a plausible assumption. For example, we would expect experts with low ability to solve a difficult item with a probability substantially larger than zero, when presented with wrong advice. <!-- SG: Maybe explain why this is not plausible. FP: like this?-->

@@

With all these considerations addressed, we are finally ready to perform a power analysis. Wrapping the `simulate` function already constructed earlier, the helper function `sim_and_analyse` performs all previous steps (simulate a dataset, fit a GLMM, compute p-values) in a single command.

@@

```{r, message=FALSE, eval=FALSE}
library(future)
plan("multisession", workers = 6)
set.seed(2)
```

The next code chunk specifies a simulation grid with different settings for both the number of subjects (`n_subjects`) and the number of items (`n_items`), each combination being repeated `rep` times. We chose 300 repetitions for the data simulation at hand as it strikes a balance between achieving a robust statistical estimate and remaining computationally feasible.
With the current settings, this simulation takes about X minutes on a MacBook Pro from 2020 with M1 chip and 16 GB working memory. If you want to quickly experiment with the code yourself, a setting with `workers = 4` and `rep = 5` should finish in less than 5 minutes, even on smaller machines.

@@

The result of the computation is a data.frame that contains the p-values of all tested hypotheses for each simulated dataset.
In some iterations (predominantly in conditions with small sample sizes), model estimation did not converge with the lme4 package. We do not remove these results because non-convergence can also happen when analyzing the real data we plan to collect, thus, we want to factor in this possibility to keep our simulation more realistic.

@@

[^3]: This code was inspired by the "Mixed Design Simulation" vignette of the faux package at <https://debruine.github.io/faux/articles/sim_mixed.html>.

```{r, eval=FALSE}
library(binom)
alpha <- 0.05
power <- sim_design %>%
  group_by(n_subjects, n_items) %>% 
  summarise(power = mean(p_H01 < alpha & p_H02 < alpha & 
                           p_H03 < alpha & p_H04 < alpha), 
    n_sig = sum(p_H01 < alpha & p_H02 < alpha & 
                  p_H03 < alpha & p_H04 < alpha),
    n = n(),
    ci.lwr = binom.confint(n_sig, n, method = "wilson")$lower,
    ci.upr = binom.confint(n_sig, n, method = "wilson")$upper, 
    .groups = "drop")
power %>%
  mutate(across(c(n_subjects, n_items), factor)) %>%
  ggplot(aes(n_subjects, n_items, fill = power)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                power, ci.lwr, ci.upr)), 
    color = "white", size = 4.5) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  xlab("number of subjects") + ylab("number of items")
```
